\chapter{Conclusion}
\label{chap:Conclusion}

\vspace{-2cm}

Here are the main conclusions of this thesis:

\begin{itemize}

\item We have studied two alterations of the public goods game with punishment. We concluded that when investing among peers, unstability in the returns, characterized by an oscillation on the enhancement factor, hinders cooperation. Moreover, the delayed punishment of defectors also makes defectors thrive.


\item We have analyzed the complexity in spacial evolutionary games, even though the system is globally stable since no significant variations of strategy frequencies occur, the local dynamics can be unstable. We obtained a method to measure the complexity of this interactions through the study of the divergence of the Hamming distance of two initially close configurations. Thanks to this tool we arrived at the typical system's time for the two configurations to be far apart.

\item We correctly classified all elementary cellular automata with the Hamming distance measure of diverging configurations. The first two classes are stable to local variations while Class-$3$ and $4$ are unstable. The distance in Class-$3$ behaves chaotically while Class-$4$ presents transitory chaos underlying the phenomenon of edge of chaos.

\item We found a new application to partial control in order to expel trajectories from a transient chaotic region. We have been able to do so with two different approaches. The first one, by accelerating the escape for the trajectories to be expelled in the least time possible. Second, by controlling the escape in order to know how much time passes since we eject control until the trajectory has escaped. Furthermore, we controlled trajectories in multistable systems by setting the time the orbit stays at each region, instead of chaotically shifting to one another.

\item Through the control partial analysis we obtained a useful tool in game theory. We designed a game of control and survival between two confronted players. With the partial control tool helps players can make decisions that gives them the victory at certain initial conditions. We analyzed different cases regarding how much information each player has.

\end{itemize}


